Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies

### 2019-05-21 man/ms.Rmd
update

* The studied relict populations showed expressions of management, likely reflecting the origin of the current forest structure. Both resilience indices of primary and secondary growth expressed that trees showed high resilience both to recent drought events and also to long-term climatic changes (*i.e.* warming), despite the ‘a priori’ vulnerability that we were expecting them to express in respone climate change on relict stands resulting from intensive land-use in history at their current rear-edge. Furthermore, all stands show a positive growth trend (stronger in high elevation stand) since the end of the 1970s matching the simultaneous increase in temperature observed.

@@

The forests of this species reaches one of their southernmost European limit in Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations between 860 m and 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August). There are eight Pyrenean oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys. Sierra Nevada mountain is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010]. In this mountain region, *Q.* *pyrenaica* woodlands represent a rear edge of their habitat distribution, which is important for determining habitat responses to expected climate change [@Hampe2005]. They are considered relictic forests [@Melendo2000; @Vivero2000], and the richest vascular-plant forest formation of Sierra Nevada [@PerezLuqueetal2014P], sheltering several endemic and endangered plant species [@Blanca1998; @Lorite2009]. These forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017]. Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" [@Vivero2000] and they are expected to suffer the impact of climate change, reducting its suitable habitats in the rear-edge due to their greater hydric requirements compared to other more drought-tolerant Mediterranean evergreen oak and pine species [@GeaIzquierdo2013].

@@

For each of the three resilience indices studied, robust two-way ANOVAs were used to test for differences between drought events (2005 and 2012) and the two studied oak populations (northern and southern aspects). These tests were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012]. When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field2012; @Wilcox2012]. Post-hoc differences were assessed pairwise using a similar boostrap test. All the robust ANOVA and post-hoc tests were carried out using the `WRS2` [@Mair2017] and `rcompanion` [@Mangiafico2017] R packages. The level of significance was set to 0.05 and adjusted for multiple comparisons. All analyses were carried out in R software [@R2017] using several packages (see Appendix S8).

@@

Greenness and tree-growth of *Q. pyrenaica* showed different sensitivity to severe droughts. During the last drought events EVI was reduced to 90.25 % (Table 1), whereas BAI was reduced to 70.83 % (Table S2), suggesting a lower sensitivity of the greenness than tree-growth to drought, particularly for sites under drier climate. For our driest site (SJ, the northern one), BAI were reduced to 44.5 % respect to that of the preceding period during 2005 drought event, whereas the EVI was reduced to 81.9 % (Tables S1 and S2). These findings are in accordance with previously works which showed that tree-growth are more sensitive metrics of forest resilience than net primary productivty [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018]. Besides the different scales of both approaches, *i.e.* tree-based *versus* pixel-based; the growth reduction seems to be more mediated by sink that by source limitations [@Gazol2018]. Response of greenness to drought not only reflects response of tree, because remote sensing indices (EVI) captures signals from all vegetation covered by the pixel (not only tree but also understory vegetation) [see @Dionisio2012]. On the other hand, tree-ring width data provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].

@@

Our study sites have been subjected to an intense exploitation of forest resources. Although the estimated age for our sampled trees was not more than 180 years (Tables 1, 2), several documents reported the presence of oaks before. For instance, the inventories of trees made by the Spanish Navy during the second half of 18th century, recorded the quantity of trees, dividing them into three categories: *new*, *growing* and *old* [@Wing2015]. For Cáñar site more than two millions of trees were reported, most of them *news*, and no *old* trees were counted, suggesting recent wood fellings. Less quantity were reported for San Juan location (circa 700 000 trees), which 220 *old* trees and 56 700 *growing* [@Cruz1991]. These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@ValbuenaCarabana2010; @Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g. @GeaIzquierdo2014; @GeaIzquierdo2015; @Dorado2017].

@@

Conclussions

Severe drought events have provoked reductions in primary and secondary growth *Q. pyrenaica* populations located in the rear-edge of their distribution.

Despite the expected vulnerability for those relict stands, we observed positive trends in primary growth (since 2000) and secondary growth (since 1970) of *Q. pyrenaica* in Sierra Nevada mountain range. Surprisingly we also observed

Those trends could have relation with the

A corto plazo, la tendencia positiva observada tanto para el crecimiento primario como para el secundario podría tener relación con varios factores. Por un lado

La tendencia creciente del crecimiento en el crecimiento observada en los últimos años podría relacionarse con un abandono de las practicas tradicionales en esta region de montaña, así como 
el abandono de las prácticas tradicionales así como por el caracter de re

El abandono de las masas de Q. pyrenaica ha provocado un estado general de degradación, que puede tener consecu

La reducción de la presión antrópica durante las últimas décadas ha provocado que estas bosques presenten un avanzado estado de degradación

>>>>

* The studied relict populations showed expressions of management, likely reflecting the origin of the current forest structure. Both resilience indices of primary and secondary growth expressed that trees showed high resilience both to recent drought events and also to long-term climatic changes (*i.e.* warming), despite the ‘a priori’ vulnerability that we were expecting them to express in respone climate change on relict stands resulting from intensive land-use in history at their current rear-edge. Furthermore, we found a positive growth trend (stronger in high elevation stand) since the end of the 1970s matching the simultaneous increase in temperature observed.

@@

The forests of this species reaches one of their southernmost European limit in Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations between 860 m and 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August). There are eight Pyrenean oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys. Sierra Nevada mountain is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010]. In this mountain region, *Q.* *pyrenaica* woodlands represent a rear edge of their habitat distribution, which is important for determining habitat responses to expected climate change [@Hampe2005]. They are considered relictic forests [@Melendo2000; @Vivero2000], and the richest vascular-plant forest formation of Sierra Nevada [@PerezLuqueetal2014P], sheltering several endemic and endangered plant species [@Blancaetal1998BC; @Loriteetal2008ABG]. These forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017]. Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" [@Vivero2000] and they are expected to suffer the impact of climate change, reducting its suitable habitats in the rear-edge due to their greater hydric requirements compared to other more drought-tolerant Mediterranean evergreen oak and pine species [@GeaIzquierdo2013].

@@

For each of the three resilience indices studied, robust two-way ANOVAs were used to test for differences between drought events (2005 and 2012) and the two studied oak populations (northern and southern aspects). These tests were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012]. When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field2012; @Wilcox2012]. Post-hoc differences were assessed pairwise using a similar boostrap test. All the robust ANOVA and post-hoc tests were carried out using the `WRS2` [@Mair2017] and `rcompanion` [@Mangiafico2017] R packages. The level of significance was set to 0.05 and adjusted for multiple comparisons. All analyses were carried out in R software [@R2017].

@@

Greenness and tree-growth of *Q. pyrenaica* showed different sensitivity to severe droughts. During the last drought events EVI was reduced to 90.25 % (Table 1), whereas BAI was reduced to 70.83 % (Table S2), suggesting a lower sensitivity of the greenness than tree-growth to drought, particularly for sites under drier climate. For our driest site (SJ, the northern one), BAI were reduced to 44.5 % respect to that of the preceding period during 2005 drought event, whereas the EVI was reduced to 81.9 % (Tables S1 and S2). These findings are in accordance with previously works which showed that tree-growth are more sensitive metrics of forest resilience than net primary productivty [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018]. Besides the different scales of both approaches, *i.e.* tree-based *versus* pixel-based; the growth reduction seems to be more mediated by sink that by source limitations [@Gazol2018]. Response of greenness to drought not only reflects response of tree, because remote sensing indices (EVI) captures signals from all vegetation covered by the pixel (not only tree but also understory vegetation) [see @Dionisioetal2012IPoGEC]. On the other hand, tree-ring width data provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].

@@

Our study sites have been subjected to an intense exploitation of forest resources. Although the estimated age for our sampled trees was not more than 180 years (Tables 1, 2), several documents reported the presence of oaks before. For instance, the inventories of trees made by the Spanish Navy during the second half of 18th century, recorded the quantity of trees, dividing them into three categories: *new*, *growing* and *old* [@Wing2015]. For Cáñar site more than two millions of trees were reported, most of them *news*, and no *old* trees were counted, suggesting recent wood fellings. Less quantity were reported for San Juan location (circa 700 000 trees), which 220 *old* trees and 56 700 *growing* [@Cruz1991]. These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@ValbuenaCarabana2010; @Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g. @GeaIzquierdo2014; @Gea-Izquierdo2015; @Dorado2017].

@@

# Conclussions

Using both dendroecological estimates of growth and MODIS satellite images as proxies of secondary and primary growth respectively, we analyzed forest resilience to severe droughts of *Q. pyrenaica* forests located at rear edge of their distribution. Severe drought events have provoked important reductions in primary and secondary growth of *Q. pyrenaica* forests. However, despite the expected vulnerability for those relict stands, we observed positive trends in primary growth (since 2000) and secondary growth (since 1970) of *Q. pyrenaica* in Sierra Nevada mountain range. Surprisingly we observed high values of resilience to drought, particularly at long-term scales. This results are consistent with research showing high genetic resilience values [@Valbuena2013; @Valbuena2017], and overall suggest this mountain region still acting as refuge for deciduous *Quercus* to climate changes [@Brewer2002; @Olalde2002; @RodriguezSanchez2010].

The differences observed in tree-growth and resilience to drought between very close sites seem to confirm that responses to drought are site-dependent [@Babst2013; Cavin2017; @Dorado2017c]. This is particularly relevant for rear-edge populations where topographic and biophysical variablity facilitates the existence of microrefugia [@VilaCabrera2019].

The analysis of tree-growth dynamics revealed suppression and release events that are consistent with socieconomic dynamic, derived from the review of historical documents. Recently researchs have combined information from both tree-ring and remote-sensing to analyze the resilience of forest to drought [*e.g.* @Gazol2018]. However our approach, which includes also an exhaustive review of historical documents, helps to disentangle the impacts of climatic and land use for forest systems. In this sense our results highligth the importance of landuse legacies for highly transformed Mediterranean systemas [@Navarro2013].

Growth projections have forecasted a decrease in productivity for *Q. pyrenaica* that would increase vulnerability of this species to climate warming at the dry edges locations [@GeaIzquierdo2013]. However, the positive trens for both primary (*i.e.* greenness) and secondary growth (BAI) observed ar our study site and the high resilience values to drought showed at long term scale, seem to suggest that Sierra Nevada constitutes still a refuge for some species. This is relevant for tree species with higher sensitivity to climate change, such us *Q. pyrenaica* (García-Valdes et al. 2013; Benito-Garzón et al. 2008), no solo por la conservación de la especie en sí, sino por los servicios ecosistémicos que estos bosques tan sigulares ofrecen.

En este sentido, los futuros trabajos deberan ir en ...

VAS POR AQUI

En estas zonas refugio en particular parece tener mucha importancia el abandono de las actividades tradicionales, que por un lado y de acuerdo a nuestros resultados parecen estar condicionando el crecimiento

Future work:

C5 Servicios ecosistémicos

Aumento de crecimiento 
Refugio para biodivesidad 
Pero ojo ---> estancamiento y/o mortalidad?? Necesidad de prácticas de gestión que aseguren un crecimiento viable, promoviendo la regeneración sexual (evitar el estancamiento) para asi mantener los 
La tendencia creciente del crecimiento en el crecimiento observada en los últimos años podría relacionarse con un abandono de las practicas tradicionales en esta region de montaña, así como 
el abandono de las prácticas tradicionales así como por el caracter de re

Además las diferencias en el crecimiento entre los dos sitios mas cercanos así como lo de la resiliencia, van en la línea de lo expresado por Vilá referente a lo de microrefugios, por lo que la gestión de estas masas, situadas en su rear edge, además ha de analizarse teniendo en cuenta este aspecto.

El abandono de las masas de Q. pyrenaica ha provocado un estado general de degradación, que puede tener consecu
La reducción de la presión antrópica durante las últimas décadas ha provocado que estas bosques presenten un avanzado estado de degradación
